### 2016-03-29 Manuscript/BiologyLetters/Missing_mammals-draft.tex
Dealt with referees 1 and 3 comments

\noindent{\Large \bf Assessment of available anatomical characters for linking living mammals to fossil taxa in phylogenetic analyses}
%TG: or actually, along the same lines:
%"Assessment of available anatomical characters for linking living and fossil mammals in phylogenetic analyses"
%TG: What do you think?

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\newpage 
\section{Introduction}
There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{ronquista2012}, combines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix that can then be used with the tip-dating method (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), producing a chronogram with living and fossil taxa at the tips.
A downside of this method is that it requires molecular data for living taxa and discrete morphological/anatomical data for both living and fossil taxa.
Sections of this data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may reduce the number of anatomical characters available.
Additionally, it has become less common to collect anatomical characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded anatomical characters).
Unfortunately this missing data can lead to errors in phylogenetic inference.
Simulations show that the ability of the Total Evidence method to recover the correct topology decreases when there is little overlap between coded anatomical characters in living and fossil taxa, and that the effect of missing data on topology is greatest when living taxa have few anatomical characters available \cite{GuillermeCooper}.
We might expect the total evidence method to perform poorly when there is little overlap between coded anatomical characters in living and fossil taxa.
This is because fossil taxa cannot be correctly placed within a clade of living species if none of its members have been coded for morphological characters.
Furthermore, simulations show that fossils are more likely to be placed in clades with for which more characters have been coded, regardless of whether this is the correct clade \cite{GuillermeCooper}.
% This is because (1) fossils will not be placed accurately within the correct clade if it contains no coded anatomical characters for living taxa; and (2) fossils have a higher probability of being placed within clades with more coded anatomical characters available for living taxa, regardless of whether this is the correct clade \cite{GuillermeCooper}.

The issues above highlight that it is crucial to have sufficient coded anatomical characters available for living taxa in a clade before using Total Evidence approaches.
However, it is unclear how many coded anatomical characters are actually available for living taxa, i.e. already coded from museum specimens and deposited in phylogenetic matrices accessible online, and how this data is distributed across clades.
Intuitively, most people assume this kind of data has already been collected, but empirical data suggest otherwise (e.g. in \cite{ronquista2012,slaterphylogenetic2013,beckancient2014}).
To investigate this further, we assess the number of available coded anatomical characters for living mammals to determine whether enough data exists to build reliable Total Evidence phylogenies.
We also determine whether the characters are phylogenetically overdispersed or clustered across mammalian orders.

@@

The number of living mammalian OTUs with available coded anatomical characters was surprisingly low at the species-level: only 17\%.
Only seven out of 28 orders have a high coverage of taxa with available coded anatomical characters.
This high coverage threshold of 75\% of taxa with available characters represents the minimum amount of data required before missing data has a significant effect on the topology of Total Evidence trees \cite{GuillermeCooper}.
Beyond this threshold, there is considerable displacement of wildcard taxa and decreased clade conservation \cite{GuillermeCooper}.
Therefore we expect difficulties in placing fossils at the species-level in most mammalian orders, but fewer issues at higher taxonomic levels.
This is important in practice because of the slight discrepancy between neontological and palaeontological species concepts.
While neontological species are described using morphology, genes, distribution etc.; palaeontological species can be based only on morphological/anatomical, spatial and temporal data (e.g. \cite{ni2013oldest}).
Therefore, many palaeontological studies use the genus (or monospecific genera) as their smallest OTU (e.g. \cite{ni2013oldest,O'Leary08022013}), so data availability at the genus-level in living mammals should be our primary concern when building phylogenies of living and fossil taxa.

@@

We acknowledge, however, that our analysis did not include all the matrices containing anatomical characters ever published.
In fact, our data collection procedure focused on including studies that provided easily accessible matrices, i.e. we did specifically not include any matrices that were only available in paper format (e.g. printed in books), non-reusable format (e.g. an image of the matrix)  or/and matrices available only upon request (e.g. by emailing the authors).
The difficulty of obtaining such matrices makes them less likely to be included in Total Evidence analyses, especially where the number of taxa to include in the phylogeny is high.

>>>>

\noindent{\Large \bf Assessment of available anatomical characters for linking living mammals to fossil taxa in phylogenetic analyses}

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\newpage 
\section{Introduction}
There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{ronquista2012}, combines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix that can then be used with the tip-dating method (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), producing a chronogram with living and fossil taxa at the tips.
A downside of this method is that it requires molecular data for living taxa and discrete morphological/anatomical data shared between both living and fossil taxa (i.e. hard tissue characters such as skeletal ones).
%
% TG: Emphasize on discrete morphological/anatomical data (often skeletal). I've added the "shared between"
%
Sections of this data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may reduce the number of anatomical characters available. %TG: should we add another "(i.e. skeletal characters)" here?
Additionally, it has become less common to collect anatomical characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded anatomical characters).
Unfortunately this missing data can lead to errors in phylogenetic inference.
% Simulations show that the ability of the Total Evidence method to recover the correct topology decreases when there is little overlap between coded anatomical characters in living and fossil taxa, and that the effect of missing data on topology is greatest when living taxa have few anatomical characters available \cite{GuillermeCooper}.
We might expect the total evidence method to perform poorly when there is little overlap between coded anatomical characters in living and fossil taxa.
This is because fossil taxa cannot be correctly placed within a clade of living species if none of its members have been coded for morphological characters.
Furthermore, simulations show that fossils are more likely to be placed in clades for which more characters have been coded, regardless of whether this is the correct clade \cite{GuillermeCooper}. %TG: That's Graham's sentence. It's not exactly what we *show* in the TEM paper but we definitely discuss this possibility in the discussion.
% This is because (1) fossils will not be placed accurately within the correct clade if it contains no coded anatomical characters for living taxa; and (2) fossils have a higher probability of being placed within clades with more coded anatomical characters available for living taxa, regardless of whether this is the correct clade \cite{GuillermeCooper}.

The issues above highlight that it is crucial to have sufficient coded hard tissue anatomical characters available for living taxa in a clade before using the Total Evidence approach.
However, it is unclear how many coded anatomical characters are actually available for living taxa, i.e. already coded from museum specimens and deposited in phylogenetic matrices accessible online, and how this data is distributed across clades.
Intuitively, most people assume this kind of data has already been collected, but empirical data suggest otherwise (e.g. in \cite{ronquista2012,slaterphylogenetic2013,beckancient2014}).
To investigate this further, we assess the number of available coded anatomical characters for living mammals to determine whether enough data exists to build reliable Total Evidence phylogenies.
We also determine whether the characters are phylogenetically overdispersed or clustered across mammalian orders.
%TG: a point to be made here maybe is that we do NOT look at hard tissue characters, rather just the presence of any characters (which is already poor enough)?
%TG: on the other hand, we might not need to emphasize that to much since, anyway, it is clearly explained in the methods (i.e. we tool all cladistic data from matrices)

@@

The number of living mammalian OTUs with available coded anatomical characters was surprisingly low at the species-level: only 17\%.
Only seven out of 28 orders have a high coverage of taxa with available coded anatomical characters.
This high coverage threshold of 75\% of taxa with available characters represents the minimum amount of data required before missing data has a significant effect on the topology of Total Evidence trees \cite{GuillermeCooper}.
Beyond this threshold, there is considerable displacement of wildcard taxa and decreased clade conservation \cite{GuillermeCooper}.
Therefore we expect difficulties in placing fossils at the species-level in most mammalian orders, but fewer issues at higher taxonomic levels.
% This is important in practice because of the slight discrepancy between neontological and palaeontological species concepts.
% While neontological species are described using morphology, genes, distribution, etc.; palaeontological species can be based only on morphological/anatomical, spatial and temporal data (e.g. \cite{ni2013oldest}).
% Therefore, many palaeontological studies use the genus (or monospecific genera) as their smallest OTU (e.g. \cite{ni2013oldest,O'Leary08022013}), so data availability at the genus-level in living mammals should be our primary concern when building phylogenies of living and fossil taxa.
%
%
%TG: Might first feeling was to just remove the last sentence but, when reading the previous one (which I like), I'm not sure if this sentence can be removed without removing the two previous ones on the species concept differences. What do you think? Anyway, it would then maybe just be better to dump the three sentences an finish this paragraph at "Therefore we expect difficulties in placing fossils at the species-level in most mammalian orders, but fewer issues at higher taxonomic levels."
%TG: On the other hand removing that gives us space for discussing the data sampling problem below
%
%

@@

We acknowledge, however, that our analysis did not include all the matrices containing anatomical characters ever published.
In fact, our data collection procedure focused on including studies that provided easily accessible matrices, i.e. we did specifically not include any matrices that were only available in paper format (e.g. printed in books), non-reusable format (e.g. an image of the matrix)  or/and matrices available only upon request (e.g. by emailing the authors).
Matrices containing anatomical characters where much more common before the advent of molecular phylogenetics and therefore are also more likely to be unavailable in a reusable format.
This might include some bias in Total Evidence analyses.
Nonetheless, these matrices are also likely to differ from more recent ones in terms of their underlying definition of homology and their coding practices (see \cite{Brazeau2011}).
Additionally, many recent morphological matrices reuse living taxa from previous matrices (see ESM1).

### 2016-03-29 Manuscript/BiologyLetters/response2.tex
Dealt with referees 1 and 3 comments

\begin{enumerate}
\item{\textcolor{blue}{line 34: spelling: ccombines = combines}}

\item{\textcolor{blue}{lines 44 - 52. I'd suggest rewording this a little bit as lines 48 - 52 are a little awkward. May I suggest the following:
"We might expect the total evidence method to perform poorly when there is little overlap between coded anatomical characters in living and fossil taxa. This is because fossil taxa cannot be correctly placed within a clade of living species if none of its members have been coded for morphological characters. Furthermore, simulations show that fossils are more likely to be placed in clades with for which more characters have been coded, regardless of whether this is the correct clade [8]."}}

\item{\textcolor{blue}{line 152: ... a carnivoran fossil is unlikely to be placed in Herpestidae because ....}}

\item{\textcolor{blue}{line 159: I'd replace "It is worth noting" with "We acknowledge"}}

\item{\textcolor{blue}{line 161: word order: provided easily accessible matrices}}
\end{enumerate}

@@

\begin{enumerate}
\item{\textcolor{blue}{The authors haven’t addressed Pete Wagner’s important comments about the use of the term “cladistics.” Morphological character data are not inherently cladistics, although they can be used for cladistics analysis, as well as Bayesian and likelihood approaches. They also haven’t done anything to clarify that the relevant character set for aligning fossil and modern data is skeletal characters. Many anatomical characters do nothing at all for the incorporation of fossil taxa.}}

@@

\item{\textcolor{blue}{I don’t agree with the assertion that the genus level should be our primary concern; more and more studies are building species-level trees with fossil taxa. I’m happy for the authors to discuss the patterns at the genus level, because it is one of the important levels, but given that genera are an artificial line in the taxonomic sand, it’s certainly not the primary one, and I’d hate to see this statement sail unopposed into the literature. I’d like to see it moderated a bit in the final version of the manuscript.}}
\end{enumerate}

@@

\begin{enumerate}
\item{\textcolor{blue}{The new title is an improvement.  However, I think that it could be altered still further: would something like: "Assessment of available anatomical characters for linking living mammals to fossil taxa in phylogenetic analyses"?  That is the main thrust of the paper, after all, and it also avoids having someone simply think: "Who cares? I have molecules" and moving on to the next article.}}

\item{\textcolor{blue}{Lines 159 - 160: "It is worth noting, however, that our analysis did not include all the matrices containing anatomical characters ever published."}}

\item{\textcolor{blue}{One thing to note is that a lot of the older analyses, which tended to be morphological, were done in the early days of modern phylogenetics, and those are the matrices that are most apt to not have pdf or html versions.  So, there might be a bias against the "pioneering" studies.  However, many of those studies will be a quarter of a century old, and the homology concepts and (especially!) coding protocols being used back then almost certainly differ from now.  Also, a lot of those matrices were "cleaned" to remove homoplasy: this was not attempted fraud, but due to the fact that computer runs that are done in hours today took weeks and weeks back then, and actually storing all of the trees that might come out of "messy" matrices was beyond the old floppy drives.  That is one reason why I rarely use matrices published before the early 1990's in my own meta-analyses.  So, if those matrices do exist and can be computerized, then I think that they should be viewed as "starting points" for coding taxa: they almost certainly will need updating.}}
\end{enumerate}

@@

\begin{enumerate}
\item{\textcolor{blue}{on L174 ("but also to any researcher focusing understanding macroevolutionary patterns and processes"), this seems to be missing an "on".}}
\end{enumerate}

>>>>

\begin{enumerate}
\item{\textcolor{blue}{line 34: spelling: ccombines = combines}}
We fixed this typo.

\item{\textcolor{blue}{lines 44 - 52. I'd suggest rewording this a little bit as lines 48 - 52 are a little awkward. May I suggest the following:
"We might expect the total evidence method to perform poorly when there is little overlap between coded anatomical characters in living and fossil taxa. This is because fossil taxa cannot be correctly placed within a clade of living species if none of its members have been coded for morphological characters. Furthermore, simulations show that fossils are more likely to be placed in clades for which more characters have been coded, regardless of whether this is the correct clade [8]."}}
We replaced the section with Referee 2's suggestion.

\item{\textcolor{blue}{line 152: ... a carnivoran fossil is unlikely to be placed in Herpestidae because ....}}
We fixed this typo.

\item{\textcolor{blue}{line 159: I'd replace "It is worth noting" with "We acknowledge"}}
We changed the start of the sentence as suggested.

\item{\textcolor{blue}{line 161: word order: provided easily accessible matrices}}
We changed the word order.

\end{enumerate}

@@

\begin{enumerate}
\item{\textcolor{blue}{The authors haven’t addressed Pete Wagner’s important comments about the use of the term “cladistics.” Morphological character data are not inherently cladistics, although they can be used for cladistics analysis, as well as Bayesian and likelihood approaches. They also haven’t done anything to clarify that the relevant character set for aligning fossil and modern data is skeletal characters. Many anatomical characters do nothing at all for the incorporation of fossil taxa.}}
We emphasised the importance of skeletal characters in the introduction as follows:
``A downside of this method is that it requires molecular data for living taxa and discrete morphological/anatomical data shared between both living and fossil taxa (i.e. hard tissue characters such as skeletal ones).''
lines @@@.

@@

We agree with this comment (and Referee 3 comment 2) and modified the paragraph discussing this caveat as follow:
``Matrices containing anatomical characters where much more common before the advent of molecular phylogenetics and therefore are also more likely to be unavailable in a reusable format.
This might include some bias in Total Evidence analyses.
Nonetheless, these matrices are also likely to differ from more recent ones in terms of their underlying definition of homology and their coding practices (see [19]).
Additionally, many recent morphological matrices reuse living taxa from previous matrices (see ESM1).''
lines @@@.
%TG: Here's how I went with the section, basically, I acknowledge referee 1's points but add Peters ones just after:
%TG: - Yes, we specifically ignored old matrices because definition of accesible data (see above)
%TG: - Yes, we do agree it could include some bias
%TG: - But these matrices are probably not so good anyway (Peter's comment)
%TG: - And anyway it doesn't solve the problem because the living taxa used in these "old" matrices are reused in new ones anyway (or at least that's what I observed in primates)
%TG: Personally I doubt the problem is due to data availability but way more to the problem we underlined in the abstract: palaeontologist don't wanna waste their time with living species (and I don't specially blame them).

\item{\textcolor{blue}{I don’t agree with the assertion that the genus level should be our primary concern; more and more studies are building species-level trees with fossil taxa. I’m happy for the authors to discuss the patterns at the genus level, because it is one of the important levels, but given that genera are an artificial line in the taxonomic sand, it’s certainly not the primary one, and I’d hate to see this statement sail unopposed into the literature. I’d like to see it moderated a bit in the final version of the manuscript.}}
Following Referee's 1 and 4's comments, we retracted this statement.

\end{enumerate}

@@

\begin{enumerate}
\item{\textcolor{blue}{The new title is an improvement.  However, I think that it could be altered still further: would something like: "Assessment of available anatomical characters for linking living mammals to fossil taxa in phylogenetic analyses"?  That is the main thrust of the paper, after all, and it also avoids having someone simply think: "Who cares? I have molecules" and moving on to the next article.}}
We changed the title as suggested by Referee 3.

\item{\textcolor{blue}{Lines 159 - 160: "It is worth noting, however, that our analysis did not include all the matrices containing anatomical characters ever published."
One thing to note is that a lot of the older analyses, which tended to be morphological, were done in the early days of modern phylogenetics, and those are the matrices that are most apt to not have pdf or html versions.  So, there might be a bias against the "pioneering" studies.  However, many of those studies will be a quarter of a century old, and the homology concepts and (especially!) coding protocols being used back then almost certainly differ from now.  Also, a lot of those matrices were "cleaned" to remove homoplasy: this was not attempted fraud, but due to the fact that computer runs that are done in hours today took weeks and weeks back then, and actually storing all of the trees that might come out of "messy" matrices was beyond the old floppy drives.  That is one reason why I rarely use matrices published before the early 1990's in my own meta-analyses.  So, if those matrices do exist and can be computerized, then I think that they should be viewed as "starting points" for coding taxa: they almost certainly will need updating.}}

% TO DO (see Referee 1 comment 2)

\end{enumerate}

@@

\begin{enumerate}

\item{\textcolor{blue}{In particular I agree with Referee 1 regarding the continued use of "cladistic" and the genus-level issue (many palaeontologists are now focused on the species-level and certainly many mammalian genera are not monotypic - this is more of an issue with dinosaur workers), and Referee 3 (Pete Wagner's) title suggestion is - I think - more likely to attract readers.}}
We dealt with these suggestions (see respectively Referee 1 comment 1, Referee 1 comment 3 and Referee 3 comment 1).

\item{\textcolor{blue}{on L174 ("but also to any researcher focusing understanding macroevolutionary patterns and processes"), this seems to be missing an "on".}}
We fixed this typo.
\end{enumerate}
